% !TeX root = ../main.tex
% Add the above to each chapter to make compiling the PDF easier in some editors.

\chapter{Background}\label{chapter:background}

Since WebAssembly on embedded devices is a new topic that just recently surfaced, we will introduce some concepts that are important to follow the thesis. While both microcontrollers and WebAssembly are widely available, there are specific details about them that are essential to be aware of in order to assess the use of WebAssembly on the ESP32 microcontroller.

\section{Microcontrollers}
A microcontroller (\textbf{MCU} for microcontroller unit) is a small computer, meant to fulfill a particular requirement without a complex operating system. They are designed for embedded applications from implantable medical devices to toys and very prominently in IoT devices. Bigger devices will often have multiple microcontrollers, each responsible for a particular function. A car, for example, could include, amongst others, an MCU to control the mirror adjustments, one to handle fuel injection and another one for traction control.

\begin{figure}[h]
    \begin{tikzpicture}
        \begin{axis}[
            x tick label style={
                /pgf/number format/1000 sep=},
            ylabel=Shipments,
            xlabel=Year,
            width=14cm,
            height=5cm,
            enlargelimits=0.05,
            ybar interval=0.7,
        ]
        \addplot coordinates {(2015,22.06) (2016,21.17) (2017,25.8) (2018,28.1) (2019,26.9) (2020,28.5) (2021,31) (2022,33.5) (2023,38) (2024,0)};
        \end{axis}
    \end{tikzpicture}
    \caption{MCU shipments worldwide from 2015 to 2023 (in billions) \autocite{ic_insights_microcontroller_2019}}
\end{figure}

Core elements of an MCU are the processor, memory, and I/O peripherals. The Processor (CPU) can be thought of as the brain of the MCU. It performs basic arithmetic, logic, and I/O operations. Memory is where any data is stored the processor needs to fulfill its tasks. Lastly, the I/O peripherals are the controller's connection to the outside world; they allow the receiving and sending of information, such as receiving a signal from a switch and turning on a light in response.

\subsection{ESP32}
For this thesis, we are not looking at big devices using multiple MCUs but instead at the ESP32 system on a chip (SoC) specifically. The ESP32 is a very popular low-cost, low power series of microcontrollers with integrated WiFi and Bluetooth. Developed by the Shanghai-based company Espressif Systems this successor to the ESP8266 offers a great platform for IoT and embedded projects \autocite{noauthor_esp32_nodate}. Compared to many other MCUs ans it's predecessor, the ESP32 has the additional processing power and I/O options that make it a great platform for developing secure IoT devices. The addional resources allow the use of more complex encryption without impacting the performance too much. It gained popularity fast after being released in September of 2016 \autocite{espressif_systems_espressif_2016}.

The ESP32 systems family provides an excellent base for many IoT applications. There are multiple versions available from ones very well suited for hobbyists to ones used for industrial manufacturers. With a low price point and small area footprint, they still provide significant performance and many operational features \autocite{maier_comparative_2017}. Examples for the ESP32 being used in IoT products include the light braceletss given at Alibaba's annual meeting in 2017 \autocite{noauthor_alibabas_2017} and DingTalks's biometric attendance monitor M1 \autocite{noauthor_dingtalks_2017}.

To reach price and power consumption targets, the ESP32 has significant hardware limitations. This introduces some constraints when working with the platform, such as 4MB flash memory and 520KB RAM installed. Compared to popular desktop operating systems like Windows and Linux, the used FreeRTOS is much more specialized. This will be explained further in the following passage.

\subsection{FreeRTOS}\label{subsec:freertos}

Many MCUs are used in applications where throughput is less important than a guaranteed performance. This is why the ESP32 uses FreeRTOS, a real-time operating system (RTOS) specifically intended to be used in time-critical situations. A key characteristic of such an operating system is the predictable behavior of the scheduler, the part of the operating system that decides which task should be run by the CPU at any given time. Most real time schedulers allow the user to set priorities for tasks in order to decide which task should be run next.

FreeRTOS is the leading RTOS amongst MCUs and is designed to be small enough to run on a microcontroller \autocite{noauthor_freertos_nodate}. Since most applications in which MCUs are used do not warrant the use of a full RTOS, FreeRTOS only provides the core scheduling functionality, timing, and synchronization primitives. This allows basic process management and synchroniszation. It can, however, be extended by using addon components, for example, to make use of a specific networking stack. FreeRTOS also built a significant developer community and support for many platforms in its 15-year development.

In more recent history, Amazon has taken over stewardship of FreeRTOs and also offers their own extension a:FreeRTOS \autocite{lardinois_amazon_nodate}. This version additionally comes with some direct integration into Amazon's AWS service \autocite{noauthor_freertos_nodate-1}. It is supposed to make the development of new IoT devices easier, especially when using Amazon's platform for the server-side processing. The core of FreeRTOS remains open source.

\section{WebAssembly}\label{sec:b_wasm}

Beginning with static HTML pages, the web has since developed into a universal application platform, accessible from many different devices running various operating systems. JavaScript is the only natively supported language on the web. However, even though it is universally used and made impressive progress in modern implementations, it still has some problems as a compilation target. The main issue being that JavaScript performance heavily depends on optimizations made by the browser, making it somewhat unpretictable and unreliable. WebAssembly addresses these issues and provides a compilation target for the web \autocite{mdn_contributors_webassembly_nodate}.

WebAssembly (WASM) was first announced in June 2015 \autocite{wagner_webassembly_2015} and reached a cross-browser consensus in March 2017 \autocite{wagner_webassembly_2017-1}. Its goal was to provide near-native performance for browser-based applications, which could only be written in JavaScript for a long time. Since reaching consensus on 2017, borser adoption makes it currently usable for about 90\% of global internet users \autocite{deveria_can_nodate}. More recently, the interest picked up around usage outside of the browser, which is also the primary concern of this thesis.

% In a study conducted on the use of WebAssembly in June of 2019, researchers looked at the top 1 million Alexa websites and found that WebAssembly was used in 1 of 600 websites \autocite{musch_new_2019}. The most common use-cases identified were malicious. Websites use it to obfuscate additional code and mine cryptocurrencies on the machines of visitors.

Being designed for the web, WebAssembly was devloped with specific goals in mind that give the format some unique attributes. It has to be safe since, on the web, code is loaded mainly from untrusted sources. It has to be fast as the primary motivation to introduce WebAssembly was to provide a compile target on the web with reliable performance. Other than the usual low-level code such as regular assembly, WebAssembly has to be portable and work in all the different environments the web is currently used in. Lastly, because the code is transmitted over the network, it has to be as small as possible to reduce bandwidth and improve latency \autocite{rossberg_bringing_2018}.

WASM is a bytecode format, designed to be a portable target for high-level languages like C/C++ or Rust. It is executed on a stack-based virtual machine on which it executes in near-native speed due to its low-level design. Still, it runs in a memory-safe environment inside the browser and is subject to the same security policies as JavaScript code would be. WebAssembly modules are loaded with the application and provide bindings to JavaScript that make them usable in the browser. They use exports to provide functions that can be called from the JavaScript context, integrating the module into web applications.

Together with the binary format of WebAssembly, there is a text format that defines a programming language with syntax and structure. Every WASM binary is a self-contained module with functions, globals, tables, imports, and exports. This concept provides both encapsulation and sandboxing since modules can only interact with their environment using imports, and the client can only access the specified exports. Inside the module, the code is organized in functions that can call each other even recursively.

WebAssembly is designed to be run on a stack machine, meaning that most of it's instructioons operate on a pushdown stack instead of registers. This leads to compact code as the location of parameters does not have to be specified, they are always expected to be on the stack. This model was chosen for WebAssembly due to it's efficient encoding, compilation and interpretation \autocite{noauthor_design_nodate}.

Other than most stack machines, WebAssembly provides structured control flow instead of arbitrary jumps \autocite{noauthor_instructions_nodate}. This limitation simplifies one-pass verification and manipulation of WebAssembly code by other tools. Figure \ref{fig:control_flow} shows the control flow syntax in WebAssembly, which shows that all block, loop, and if constructs must be terminated with the end instruction and have to be correctly nested. They bracket nested sequences of instructions referred to as blocks in WASM; each block has an implicit label, which is the target for branch instructions.

\begin{figure}[h]
    \begin{equation}\nonumber
        \begin{split}
            \begin{array}{llcl}
                \mathit{instr} &::=&
                \dots \\&&|&
                \mathsf{nop} \\&&|&
                \mathsf{unreachable} \\&&|&
                \mathsf{block}~\mathit{resulttype}~\mathit{instr}^\ast~\mathsf{end} \\&&|&
                \mathsf{loop}~\mathit{resulttype}~\mathit{instr}^\ast~\mathsf{end} \\&&|&
                \mathsf{if}~\mathit{resulttype}~\mathit{instr}^\ast~\mathsf{else}~\mathit{instr}^\ast~\mathsf{end} \\&&|&
                \mathsf{br}~\mathit{labelidx} \\&&|&
                \mathsf{br\_if}~\mathit{labelidx} \\&&|&
                \mathsf{br\_table}~\mathit{vec}(\mathit{labelidx})~\mathit{labelidx} \\&&|&
                \mathsf{return} \\&&|&
                \mathsf{call}~\mathit{funcidx} \\&&|&
                \mathsf{call\_indirect}~\mathit{typeidx} \\
            \end{array}
        \end{split}
    \end{equation}
    \caption{WebAssembly control flow syntax}
    \label{fig:control_flow}
\end{figure}

Since they are used later on, we want to explain the \lstinline{br_if}, \lstinline{return} and \lstinline{call} instructions. The first two are branch instructions that will reference a specific block label. \lstinline{br_if} performs a conditional branch and is used, for example, to execute if statements. \lstinline{return} is a special unconditional branch that always branches to the outermost block. The \lstinline{call} instruction invokes another function by taking the necessary arguments from the stack and returning the result of the execution of the function.

The WebAssembly runtime keeps all the global state that can be manipulated by the module in the store. As shown in Figure \ref{fig:store_content} the store consists of all instances of functions, tables, memories and globals that have been allocated while the abstract machine was running. Besides the store, almost all instructions interact with an implicit stack. This stack contains values, labels, and call frames of active function calls. Thus, if a function returns a result, it will be the first entry on the stack after the execution has finished \autocite{noauthor_runtime_nodate}.

\begin{figure}[h]
    \begin{equation}\nonumber
        \begin{split}
            \begin{array}{llll}
            \mathit{store} &::=& \{~
            \begin{array}[t]{l@{~}ll}
                \mathsf{funcs} & \mathit{funcinst}^\ast, \\
                \mathsf{tables} & \mathit{tableinst}^\ast, \\
                \mathsf{mems} & \mathit{meminst}^\ast, \\
                \mathsf{globals} & \mathit{globalinst}^\ast ~\} \\
            \end{array}
            \end{array}
        \end{split}
    \end{equation}
    \caption{WebAssembly store contents}
    \label{fig:store_content}
\end{figure}

\subsection{WebAssembly for IoT}

While WASM was developed for the web, it carefully avoids any dependencies on the web. It is meant to be an open standard that can be embedded in a variety of ways \autocite{noauthor_non-web_nodate}. The goals mentioned above, which WebAssembly achieves, make it an exciting format to explore on embedded devices. Due to its aim to be universal, it would allow the use of languages on MCUs that were not previously supported, and since it is already meant to be transmitted over the network, over the air updates of code running on the controller are also possible. To achieve portability, the source level interface libraries would have to map the host environments' capabilities either at build time or runtime.

While WebAssembly in an IoT context is an up-and-coming concept, the development has only just begun. The best support for WASM right now is in the browser, of course, but out of browser runtimes and compilers are being devloped, implemented in various languages such as Python, JavaScript, C, Rust and even OCaml \autocite{akinyemi_appcypherawesome-wasm-runtimes_2020}. Runtimes meant to be used on MCUs are much rarer and not as mature yet. Given the significant interest in the idea though, and the working groups' avoidance of web dependencies, it can be assumed that this situation will change in the future.

\subsection{Interpreters}

With our specific use-case in mind, the WASM3 \autocite{noauthor_wasm3wasm3_2020} engine was chosen because one of its main goals is to run WebAssembly on MCUs. Contrary to many other engines, it does not follow a just-in-time (JIT) compilation pattern though, but instead acts as an interpreter.

Interpreters are computer programs that execute a program. They pose a different concept to compiled execution, where a program would be translated to machine code before being run directly on the CPU. While offering multiple advantages, the main drawback is the execution speed compared to native code execution, which is often slower by order of magnitude and sometimes more. The overhead is generated by the interpreter having to analyze the program code before it can be executed.

Interpreters thus offer benefits in development speed since the code does not have to be recompiled in order to run and in portability because the same code could be run on multiple platform-specific interpreters without the need to compile it into the native machine code of multiple platforms. For our use-case, the interpreter executes WASM instructions, allowing the dynamic loading of modules and running them in the chosen environment.

\section{Microbenchmarking}

Benchmarking is any form of measurement used to qualify the behavior of a system. The most obvious examples would be measuring performance, energy, or memory consumption, but also reliability and temperature stability could contribute to a benchmark. Building useful benchmarks is hard because a program has to be created that yields repeatable and consistent results. A notable effort in the world of microcontroller benchmarking is the embedded microprocessor benchmark consortium (EEMBC) \autocite{noauthor_embedded_nodate}.

The EEMBC is an industry association that has been designing benchmarks for over 20 years. The consortium offers multiple benchmarks, all meant to cover specific use-cases of embedded controllers from ultra-low-power IoT applications, over processor performance measurements to a recent benchmark designed to assess machine learning performance \autocite{noauthor_benchmark_nodate}.

\section{Summary}
Microcontrollers are tiny, and restricted computers meant to perform a specific task. They are part of our everyday life and are used in a great variety of applications. The ESP32 system on a chip is a prevalent family of Microcontrollers that are very well suited for use in the Internet of Things devices. They are affordable but still offer WiFi and Bluetooth connectivity.

WebAssembly is a new bytecode format that was designed for browsers, with the aim of allowing developers to run optimized low-level code at near-native speeds. It is supported by all major browsers and in active development. However, people are not only interested in WebAssembly on the web itself but also recognize its potential as a new universal and portable bytecode outside of the browser.